Liveness-Based Garbage Collection for Lazy Languages

### 2016-02-03 paper/paper.tex
Changed order of statements in introduction

\todelete{While  this  idea has  been  shown  to  be effective  for  a
  first-order     {\em     eager}    language~\cite{asati14lgc},     a
  straightforward extension of the technique  is not possible for lazy
  languages,  where   heap-allocated  objects  may   include  closures
  (runtime   representations    of   unevaluated    expressions).    }
\added{Designing a liveness based garbage collector for lazy languages
  poses significant challenges.}  Firstly, since  data is made live by
evaluation of closures, and in lazy languages the place in the program
where  this evaluation  takes place  cannot be  statically determined,
laziness   complicates  liveness   analysis  itself.    Moreover,  for
liveness-based garbage collection  to be effective, we  need to extend
it to closures apart from evaluated  data.  Since a closure can escape
the  scope in  which it  was  created, it  has to  carry the  liveness
information  of  its  free  variables.  As  execution  progresses  and
possible  future   uses  are   eliminated,  we  update   the  liveness
information  in a  closure with  a more  precise version.   Because of
these reasons,  the garbage collector also  becomes significantly more
complicated than a liveness-based collector for a eager language.

Experiments    with   a    single    generation   copying    collector
(Section~\ref{sec:experiments})   confirm  the   expected  performance
benefits.  Liveness-based collection results in an increase in garbage
reclaimed.  As  a consequence, there is  a reduction in the  number of
collections from 1.6X  to 23X and a decrease  in the minimum
memory requirement from  1X to 1389X.  As  an added benefit,
there is also a reduction in the overall execution time in 5 out of 12
benchmark programs.

@@

In  this work  we show  that  static analysis  of heap  data can  help
garbage  collectors   in  reclaiming   more  garbage.    The  specific
contributions of this paper are
\added{
\begin{myitemize}
\item We propose a liveness based garbage collection scheme for a lazy
  first order functional  language and prove its  correctness.  To the
  best  of our  knowledge, this  is the  first work  on liveness-based
  garbage    collection   for    lazy    languages.    All    previous
  attempts~\cite{shaham01heap,  ran.shaham-sas03,  shaham02estimating,
    asati14lgc,    karkare06effectiveness}   have    involved   either
  imperative  or  eager  functional  languages.   The  novely  of  our
  approach is in  handling of  liveness of  functional values  (i.e.,
  closures), an issue that arises due to laziness.

\item  We  formulate a  liveness  analysis  for the  lazy  first-order
  functional language  and prove  its correctness. The  proof involves
  specifying  liveness   for  the  language  through   a  non-standard
  semantics and then proving the  analysis correct with respect to the
  specification.

\item The  analysis results  in a set  of context-free  grammars along
  with  a fixed  set  of non-context-free  productions.  The  decision
  whether to  copy a  cell during garbage  collection translates  to a
  membership problem for such grammars.   In this paper, we have shown
  the undecidability  of this  problem. We  circumvent the  problem by
  over-approximating the grammars with DFAs.

\item We have implemented a garbage  collector that uses the result of
  liveness  analysis to  retain  live cells.   Our experiments  reveal
  interesting  space-time   trade-offs  in  the  engineering   of  the
  collector---for  example, updating  liveness information  carried in
  closures during  execution results in more  garbage being collected.
  Empirical results  show the effectiveness of  liveness-based garbage
  collection.
\end{myitemize}
}
\todelete{
\begin{myitemize}
\item  We  formulate  a  liveness  analysis for  a  lazy  first-order
  functional language and  prove its correctness. The 
  proof involves specifying liveness through a non-standard semantics  and then proving the analysis 
  correct with respect to the specification.   
\item The analysis results  in a set  of context-free  grammars along
  with  a fixed  set  of non-context-free  productions.  The  decision
  whether   to  copy  a  cell   during  garbage  collection
  translates  to a  membership problem  for such  grammars.  In  this
  paper, we  have shown the  undecidability of this problem and 
  circumvent the  problem by over-approximating the  grammars
  with DFAs.
\item We  have implemented a garbage  collector that uses the  DFAs to
  retain live  cells.  Our  experiments reveal  interesting space-time
  trade-offs  in  the  engineering  of  the  collector---for  example,
  updating liveness  information carried in closures  during execution
  results  in   more  garbage  being  collected.    Empirical  results
  show the effectiveness of liveness-based garbage collection.
\end{myitemize}

>>>>

\todelete{While this idea has been shown  to be effective for a first-order {\em
  eager}  language~\cite{asati14lgc}, a  straightforward extension  of
the technique is not possible for lazy languages, where heap-allocated
objects may  include closures (runtime representations  of unevaluated
expressions).}
We encounter significant challenges in defining a liveness based garbage 
collector for lazy languages. 
\added{
Firstly, during garbage collection the heap contains both fully 
evaluated data and unevaluated thunks/closures. 
For liveness-based
garbage collection to  be effective, we need to extend  it to closures
apart from  evaluated data. 
Moreover, since  data  is made  live  by evaluation  of
closures, and  in lazy languages the  place in the program  where this
evaluation  takes  place  cannot be  statically  determined,  laziness
complicates  liveness analysis  itself.
Since closures can escape the scope in which 
they were created, they need to carry liveness information of its free 
variables. 
As  execution progresses and possible  future uses are
eliminated, we  update the  liveness information in  a closure  with a
more precise version.  Because of these reasons, the garbage collector
also  becomes significantly  more  complicated  than a  liveness-based
collector for a eager language.
}

\todelete{
Firstly, since  data  is made  live  by evaluation  of
closures, and  in lazy languages the  place in the program  where this
evaluation  takes  place  cannot be  statically  determined,  laziness
complicates  liveness analysis  itself.  Moreover,  for liveness-based
garbage collection to  be effective, we need to extend  it to closures
apart from  evaluated data.  Since a  closure can escape the  scope in
which it was created, it has  to carry the liveness information of its
free variables. As  execution progresses and possible  future uses are
eliminated, we  update the  liveness information in  a closure  with a
more precise version.  Because of these reasons, the garbage collector
also  becomes significantly  more  complicated  than a  liveness-based
collector for a eager language.
}
Experiments    with   a    single    generation   copying    collector
(Section~\ref{sec:experiments})   confirm  the   expected  performance
benefits.  Liveness-based collection results in an increase in garbage
reclaimed.  As  a consequence, there is  a reduction in the  number of
collections from 1.6X  to 23X and a decrease  in the minimum
memory requirement from  1X to 1389X.  As  an added benefit,
there is also a reduction in the overall execution time in 5 out of 12
benchmark programs.

@@

In  this work  we show  that  static analysis  of heap  data can  help
garbage  collectors   in  reclaiming   more  garbage.    The  specific
contributions of this paper are
\added{
\begin{myitemize}
\item  We propose a liveness based garbage collection scheme for a lazy first order 
functional language and prove its correctness. 
To the best of our knowledge this is the first effort to define a
liveness based garbage collection for a lazy language. 
Our main contribution is the way we handle liveness of functional values i.e closures which 
distinguishes our work from earlier work in this area.

\item
  We  formulate  a  liveness  analysis for  the  lazy  first-order
  functional language and  prove its correctness. The 
  proof involves specifying liveness through a non-standard semantics  and then proving the analysis  
  correct with respect to the specification.

\item 
  The analysis results  in a set  of context-free  grammars along
  with  a fixed  set  of non-context-free  productions.  The  decision
  whether   to  copy  a  cell   during  garbage  collection
  translates  to a  membership problem  for such  grammars.  In  this
  paper, we  have shown the  undecidability of this problem. We 
  circumvent the  problem by over-approximating the  grammars
  with DFAs.
\item 
We  have implemented a garbage  collector that uses the  result of liveness analysis to
  retain live  cells.
Our  experiments reveal  interesting space-time
  trade-offs  in  the  engineering  of  the  collector---for  example,
  updating liveness  information carried in closures  during execution
  results  in   more  garbage  being  collected.    Empirical  results
  show the effectiveness of liveness-based garbage collection.
\end{myitemize}
}
\todelete{
\begin{myitemize}
\item  We  formulate  a  liveness  analysis for  a  lazy  first-order
  functional language and  prove its correctness. The 
  proof involves specifying liveness through a non-standard semantics  and then proving the analysis 
  correct with respect to the specification.   
\item The analysis results  in a set  of context-free  grammars along
  with  a fixed  set  of non-context-free  productions.  The  decision
  whether   to  copy  a  cell   during  garbage  collection
  translates  to a  membership problem  for such  grammars.  In  this
  paper, we  have shown the  undecidability of this problem and 
  circumvent the  problem by over-approximating the  grammars
  with DFAs.
\item We  have implemented a garbage  collector that uses the  DFAs to
  retain live  cells.  Our  experiments reveal  interesting space-time
  trade-offs  in  the  engineering  of  the  collector---for  example,
  updating liveness  information carried in closures  during execution
  results  in   more  garbage  being  collected.    Empirical  results
  show the effectiveness of liveness-based garbage collection.
\end{myitemize}
